---
title: |
  | Assignment 3 
  | Econometrics I
subtitle: "Universidad Carlos III de Madrid"
author: "Gabriel Merlo"
date:
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

# 1. Predicting wages in USA

```{r include_packages, results = "hide"}
# install (if missing) packages
list_packages <- c("dplyr", "formatR", "ggplot2", "glmnet", "hdm", "Hmisc")
new_packages <- list_packages[!(list_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

# Load packages
sapply(list_packages, require, character.only = TRUE)
```
## (a) Load, prepare and summarize the data

```{r part_a, results = "hide"}
## Load prepare and summarize the data

# Load Census data from the US for the year 2012
data("cps2012")

# Check data and summarize
str(cps2012)
summary(cps2012) #no missing obs.
head(cps2012)
describe(cps2012)
```
## (b) Apply Ridge-Regression with cross-validation (CV)

The ridge regression coefficient estimates $\hat{\beta}_R$ are the values that minimize $$\sum_{i=1}^{n}\left(y_i-\beta_0-\sum_{j=1}^{p}\beta_j x_{ij}\right)^2+\lambda\sum_{j=1}^{p}\beta_j^2$$

Where $\lambda\geq 0$ is a penalty parameter that shrinks the estimates of $\beta_j$ towards zero as $\lambda\to\infty$ (with the extreme case were all coefficient estimates are zero and the model contains no predictors). When $\lambda=0$ the ridge coefficient estimates are the same as least squares (OLS) estimates.

### (i) Apply ridge regression to the previous dataset for the the default grid of values of lambda. Plot the 10-fold CV MSE as a function of lambda.

```{r part_b_i, tidy=TRUE, tidy.opts=list(width.cutoff=80)}

# Define dependent variable y: lnw
y <- cps2012$lnw

# Define matrix of predictors
x <- cps2012 %>% select(female, widowed, divorced, separated, nevermarried, hsd08, hsd911, hsg, cg, ad, mw, so, we, exp1, exp2, exp3) %>% data.matrix()

## Apply ridge regression using glmnet

# Define range of lambda values
lambdas <- 10^seq(10, -2, by = -.1)

# Estimate ridge model (we need to set alpha = 0 to get ridge coefficients)
fit_ridge <- glmnet(x, y, alpha = 0, lambda = lambdas)
summary(fit_ridge)
```


