---
title: "Assignment 3"
author: "Ignacio"
date: "02/12/2020"
output:
  pdf_document: default
html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("hdm")
#install.packages("MXM")
#install.packages("caret")
#install.packages("dplyr")
#install.packages("glmnet")
#install.packages("MLmetrics")
#install.packages("rmarkdown")
#install.packages("geometry")

library("rmarkdown")
library("MLmetrics")
library("caret")
library(hdm)
library("MXM")
library("magrittr")
library("dplyr")
library("glmnet")
library(lmtest)
library(sandwich)
library(geometry)
data("cps2012")


```



# a)


```{r}

summary(cps2012)
```

```{r, echo=FALSE}

#Scaling the variables

cps2012_sc <- cps2012 %>% select(lnw, female, widowed, divorced, separated, nevermarried, hsd08, hsd911, hsg,
                        cg, ad, mw, so, we, exp1, exp2, exp3)  %>% data.frame()
#%>% scale(center=TRUE)

#summary(cps2012_sc)
```

#b)
```{r, echo=FALSE}

# This lines of code create a data frame which contain interaction variables with female gender that we
# will use at part d)

aux <- cps2012_sc %>% select(widowed, divorced, separated, nevermarried, hsd08, hsd911, hsg,
                                    cg, ad, mw, so, we, exp1, exp2, exp3) %>% data.matrix()

interact_matrix <- aux*cps2012_sc$female

colnames(interact_matrix) <- paste("interaction_g", 1:ncol(interact_matrix), sep="")


interact_matrix_dfr = as.data.frame(interact_matrix)
#cps2012_sc = as.data.frame(cps2012_sc)


cps2012_sc$generated_id <- 1:nrow(cps2012_sc)
interact_matrix_dfr$generated_id <- 1:nrow(interact_matrix_dfr)

cps2012_sc_int = merge(x=cps2012_sc,y=interact_matrix_dfr,by = "generated_id" )
cps2012_sc_int <- subset( cps2012_sc_int, select = -generated_id )


# Here we are constructing training data with 70% of the sample and testing data with the remaining 30%

set.seed(50)

indx = sample(1:nrow(cps2012_sc_int), 0.7*nrow(cps2012_sc_int)) 

train = cps2012_sc_int[indx,] %>% scale(center=TRUE) # Create the training data 
test = cps2012_sc_int[-indx,] %>% scale(center=TRUE) # Create the test data




summary(train)

train_dfr = as.data.frame(train)
test_dfr = as.data.frame(test)

y_train <- train_dfr %>% select(lnw) %>% data.matrix()

#matrix with predictors for the regression using training data

x_train <- train_dfr %>% select(female, widowed, divorced, separated, nevermarried, hsd08, hsd911, hsg,
                        cg, ad, mw, so, we, exp1, exp2, exp3) %>% data.matrix()

#matrix with predictors for the regression with interactions

x_int_train = train_dfr %>% select(female, widowed, divorced, separated, nevermarried, hsd08, hsd911, hsg,
                                   cg, ad, mw, so, we, exp1, exp2, exp3,
                                   interaction_g1,interaction_g2,interaction_g3,interaction_g4,interaction_g5,
                                   interaction_g6,interaction_g7,interaction_g8,interaction_g9,interaction_g10,
                                   interaction_g11,interaction_g12,interaction_g13,interaction_g14,interaction_g15
                                   ) %>% data.matrix()

x = cps2012_sc_int %>% select(female, widowed, divorced, separated, nevermarried, hsd08, hsd911, hsg,
                                   cg, ad, mw, so, we, exp1, exp2, exp3,
                                   interaction_g1,interaction_g2,interaction_g3,interaction_g4,interaction_g5,
                                   interaction_g6,interaction_g7,interaction_g8,interaction_g9,interaction_g10,
                                   interaction_g11,interaction_g12,interaction_g13,interaction_g14,interaction_g15
                                   ) %>% data.matrix()

y <- cps2012_sc_int %>% select(lnw)  %>% data.matrix()

x_train_dfr = as.data.frame(x_train)

#################################################

#testing data

y_test <- test_dfr %>% select(lnw) %>% data.matrix()
x_test <- test_dfr %>% select(female, widowed, divorced, separated, nevermarried, hsd08, hsd911, hsg,
                           cg, ad, mw, so, we, exp1, exp2, exp3) %>% data.matrix()
x_int_test = test_dfr %>% select(female, widowed, divorced, separated, nevermarried, hsd08, hsd911, hsg,
                                   cg, ad, mw, so, we, exp1, exp2, exp3,
                                   interaction_g1,interaction_g2,interaction_g3,interaction_g4,interaction_g5,
                                   interaction_g6,interaction_g7,interaction_g8,interaction_g9,interaction_g10,
                                   interaction_g11,interaction_g12,interaction_g13,interaction_g14,interaction_g15
)%>% data.matrix()







# RIDGE

## i)

# Ridge regression for the default grid of values of lambda

ridge <- cv.glmnet(x_train, y_train, alpha = 0, standardize = TRUE) 

lambda10 <- 10^seq(2, -3, by = -.1)
ridge <- cv.glmnet(x_train, y_train, alpha = 0, lambda = lambda10, nfolds = 10, standardize = TRUE) 
```
# Plot of the 10-fold CV MSE as a function of lambda

plot(ridge)

## ii)

optimal_lambda <- ridge$lambda.min
print(optimal_lambda)

fit <- ridge$glmnet.fit

summary(fit) 

#ridge prediction 

ridge <- glmnet(x_train, y_train, alpha = 0, lambda = optimal_lambda, nfolds = 10, standardize = TRUE) 

predictions_test_ridge <- predict(ridge, s = optimal_lambda, newx = x_test, newdata = test)


RMSE_ridge = sqrt(MSE(predictions_test_ridge, y_test ))


print(RMSE_ridge)

# Number of variables

ridge.coef=predict(ridge,type="coefficients",
s=optimal_lambda)

sum(ridge.coef!=0)


## iii) 

#ols 

lr = lm(y_train ~ female + widowed + divorced + separated + nevermarried + hsd08 + hsd911 + hsg +
        cg + ad + mw + so + we + exp1 + exp2 + exp3, data=train_dfr )

coeftest(lr, vcov.=vcovHC(lr, "HC0")) # Robust Standard Errors

predictions_ols = predict(lr, newx = x_test, newdata = test)

RMSE_ols = sqrt(MSE(predictions_ols, y_test ))

print(RMSE_ols)







summary(lr)





#LASSO

cv_lasso <- cv.glmnet(x_train,y_train, alpha= 1, lambda = lambda10, nfolds = 10)

#cv_lasso <- cv.glmnet(x_int_train,y_train, alpha= 1, type.measure = 'mse')

optimal_lambda_ls <- cv_lasso$lambda.min 


print(optimal_lambda_ls)
plot(cv_lasso)



predictions_test_ls <- predict(cv_lasso, s = optimal_lambda_ls, newx = x_test)
RMSE_lasso = sqrt(MSE(predictions_test_ls, y_test ))
print(RMSE_lasso)

# Number of variables

lasso.coef=predict(cv_lasso,type="coefficients",
s=optimal_lambda_ls)


sum(lasso.coef!=0)


##########

#d) #########################

#OLS

lr = lm(y_train ~ female + widowed + divorced + separated + nevermarried + hsd08 + hsd911 + hsg +
          cg + ad + mw + so + we + exp1 + exp2 + exp3 + interaction_g1 + interaction_g2 + interaction_g3
        + interaction_g4+ interaction_g5 +interaction_g6 + interaction_g7 + interaction_g8 + interaction_g9
        + interaction_g10 + interaction_g11 + interaction_g12 + interaction_g13 + interaction_g14 + 
          interaction_g15, data=train)
          
       


predictions_ols = predict(lr, newx = x_int_test, newdata = test)

RMSE_ols = sqrt(MSE(predictions_ols, y_test ))

print(RMSE_ols)

summary(lr)

#RIDGE

ridge <- glmnet(x_int_train, y_train, alpha = 0, lambda = optimal_lambda, nfolds = 10, standardize = TRUE) 

predictions_test_ridge <- predict(ridge, s = optimal_lambda, newx = x_test, newdata = test)

RMSE_ridge = sqrt(MSE(predictions_test_ridge, y_test ))


print(RMSE_ridge)

ridge.coef=predict(ridge,type="coefficients",
s=optimal_lambda)

sum(ridge.coef!=0)

#LASSO

cv_lasso <- glmnet(x_int_train,y_train, alpha= 1, lambda = optimal_lambda_ls, nfolds = 10)

predictions_test_ls <- predict(cv_lasso, s = optimal_lambda_ls, newx = x_int_test)
RMSE_lasso = sqrt(MSE(predictions_test_ls, y_test ))
print(RMSE_lasso)


# Number of variables

lasso.coef=predict(cv_lasso,type="coefficients",
s=optimal_lambda_ls)[1:20,]


sum(lasso.coef!=0)

sum(lasso.coef=0)



round(lasso.coef[lasso.coef!=0],4)


#Oaxaca descomposition

cv_lasso <- cv.glmnet(x,y, alpha= 1, lambda = lambda10, nfolds = 10)


optimal_lambda_ls <- cv_lasso$lambda.min 




x_w = df_w %>% select(female, widowed, divorced, separated, nevermarried, hsd08, hsd911, hsg,
                                   cg, ad, mw, so, we, exp1, exp2, exp3,
                                   interaction_g1,interaction_g2,interaction_g3,interaction_g4,interaction_g5,
                                   interaction_g6,interaction_g7,interaction_g8,interaction_g9,interaction_g10,
                                   interaction_g11,interaction_g12,interaction_g13,interaction_g14,interaction_g15
                                   ) %>% data.matrix()

y_w <- df_w %>% select(lnw)  %>% data.matrix()


x_m = df_m %>% select(female, widowed, divorced, separated, nevermarried, hsd08, hsd911, hsg,
                                   cg, ad, mw, so, we, exp1, exp2, exp3,
                                   interaction_g1,interaction_g2,interaction_g3,interaction_g4,interaction_g5,
                                   interaction_g6,interaction_g7,interaction_g8,interaction_g9,interaction_g10,
                                   interaction_g11,interaction_g12,interaction_g13,interaction_g14,interaction_g15
                                   ) %>% data.matrix()

y_m <- df_m %>% select(lnw)  %>% data.matrix()


cv_lasso_w <- cv.glmnet(x_w,y_w, alpha= 1, lambda = lambda10, nfolds = 10)

cv_lasso_m <- cv.glmnet(x_m,y_m, alpha= 1, lambda = lambda10, nfolds = 10)


lasso_coefw=predict(cv_lasso_w,type="coefficients",
s=optimal_lambda_ls)[1:20,]

lasso_coefm=predict(cv_lasso_m,type="coefficients",
s=optimal_lambda_ls)[1:20,]


df_w <- subset(cps2012_sc_int, cps2012_sc_int$female == 1)
df_m <- subset(cps2012_sc_int, cps2012_sc_int$female == 0)

aux_w <- list( rep(1,length(df_w$lnw)), df_w$widowed, df_w$divorced, df_w$separated, df_w$nevermarried, df_w$hsd08, df_w$hsd911, df_w$ hsg, df_w$cg, df_w$ad, df_w$ mw, df_w$so, df_w$we, df_w$exp1, df_w$exp2, df_w$exp3)

aux_m <- list(rep(1,length(df_w$lnw)), df_m$widowed, df_m$divorced, df_m$separated, df_m$nevermarried, df_m$hsd08, df_m$hsd911, df_m$ hsg, df_m$cg, df_m$ad, df_m$mw, df_m$so, df_m$we, df_m$exp1, df_m$exp2, df_m$exp3)


over <- mean(df_w$lnw) - mean(df_m$lnw)

print(over)

Xbar1<- unlist(lapply(aux_w, mean))
Xbar0<- unlist(lapply(aux_m, mean))



ols_w <- lm(lnw ~ widowed +divorced + separated + nevermarried + hsd08 + hsd911 + hsg + cg + ad + mw + so + we
                  + exp1 + exp2 +exp3, data = subset(cps2012_sc_int, cps2012_sc_int$female == 1))
                    


ols_m <- lm(lnw ~ widowed +divorced + separated + nevermarried + hsd08 + hsd911 + hsg + cg + ad + mw + so + we
                  + exp1 + exp2 +exp3, data = subset(cps2012_sc_int, cps2012_sc_int$female == 0))

str1 <- coef(ols_w) - coef(ols_m)
str <- dot(Xbar1,str1)

comp <- dot(Xbar1 - Xbar0,coef(ols_m))


